\propertitle

### 2019-12-05 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-04T20:53:39+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand*{\yRR}{R}
\newcommand*{\yR}{\bm{\yRR}}

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{P_{\text{me}}}
\newcommand*{\pxx}{p}
\newcommand*{\pxxx}{P}
\newcommand*{\Nd}{\textrm{N}}
\newcommand*{\ycmm}{C}
\newcommand*{\ycm}{\mathte{\ycmm}}
\newcommand*{\ycs}{\bm{\sigma}}
\newcommand*{\yL}{L}
%\newcommand*{\yd}{\bm{\varDelta}}
\newcommand*{\ydi}{\varDelta}

@@

\section{Sketch of inference with the probability calculus}
\label{sec:inference_prob}

The first and most important assumption of this kind of maximum-entropy
applications in neuroscience can be expressed in two equivalent ways:
\begin{enumerate*}[(\roman*)]
  % \item the order of the time bins has no relevance;
\item the frequencies of the activity during a recording are
  informationally sufficient for inferences about unrecorded times; \item
  our degree of belief about the sequence of activities is exchangeable
  \citep[\chap~4]{bernardoetal1994_r2000}{dawid2013}, that is, invariant
  under permutations of the time bins.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
\citep{definetti1930,hewittetal1955} is that the probability of observing a
sequence of activities $(\yA_{1},\dotsc,\yA_{T})$ with frequencies
$(\yFF_{\yA})$ is a mixture of terms
\begin{equation}
  \label{eq:mixture_from_definetti}
%  \exp[-T\;\sh(\yF;\ynu)] \quad
%  \pxxx[(\yA_{1},\dotsc,\yA_{T})] \propto
  \prod_{\yA=1}^{N}{\nu_{\yA}}^{T\,\yFF_{\yA}} \quad
  \text{with weights }q(\ynu),
\end{equation}
where $\ynu\defd (\nu_{\yA})$ represents the long-run frequency
distribution for the activity. % Other maximum-entropy models are also used
% in neuroscience, such as kinetic Ising models
% \cite{kawasaki1972,mezardetal2011,battistinetal2015}, based on the slightly
% less drastic assumption of invariance under permutations of sequences
% having the same starting state and the same transition counts
% \citep{diaconisetal1980c,geetal2012}.
Note that the maximum-entropy method in statistical physics is meant to be
used not with this assumption, but with averages across \emph{repetitions}
of the same experiment, because its constraints represent macroscopically
\emph{reproducible} quantities and kinematics, even in non-equilibrium
\citep{jaynes1957,guntonetal1983,tikochinskyetal1984,grandy1988,berryetal1988,deroecketal2005_r2006,touchette2009}.

The second assumption is specific to our application: all subsets of $\yA$
neurons in the larger population are equally likely to give rise to total
activity $\yA$, at every time bin. A mathematically equivalent assumption
is that $n$ neurons are sampled anew at each time bin. This assumption
allows us to factorize the joint probability of the two sequences of
activities and apply the sampling
relation~\eqref{eq:conditional_hypergeometric} of
appendix~\ref{sec:derivation_maxent_fullnet} at each bin:
\begin{equation}
  \label{eq:assume_sampled_anew}
  \pxx[(\ya_{1},\dotsc,\ya_{T}) \| (\yA_{1}, \dotsc, \yA_{T}) ]
  = \prod_{t=1}^{T} \pxx(\ya_{t} \| \yA_{t})
  = \prod_{t=1}^{T} \yG_{\ya_{t}\yA_{t}}.
\end{equation}

Both assumptions above make sense as reference hypotheses for judging the
informational relevance of correlations of some order, as in
\sect~\ref{sec:marginalization}, to be compared with hypotheses about more
realistic time dependences and clustering.

From these two assumptions the probability calculus allows us to derive not
only the probability for the frequency distribution $\yF$, but also that
for the joint frequencies of total activities of sample and larger
population, $J_{\ya,\yA}$, conditional on the observed frequency
distribution for the sample $\yf$. The result is again a mixture of terms
\begin{equation}
  \label{eq:prob_joint_frequencies}
%\pxx[\yF \| \yf ] \propto
  \prod_{\ya,\yA}\bigl(\yG_{\ya\yA}\;\nu_{\yA}\bigr)^{T\,J_{\ya,\yA}}
  \qquad
  \text{with } \tsum_{\yA}J_{\ya,\yA}=\yff_{\ya}
  %\quad \tsum_{\ya}J_{\ya,\yA}=\yFF_{\yA},
\end{equation}
with weights $q(\ynu)$.

The maximum-entropy method approximates $T$ in the probability above with
infinity, thus treating it as a delta. This leads to two approximate equalities:
\begin{enumerate*}[label=(\roman*)]
\item of the recorded and long-run frequency distributions $\yF$ and
  $\ynu$, and therefore also of their probability distributions,
\item of the hypergeometric distribution $\yG_{\ya\yA}$ and the frequency
  of the sample activity $\ya$ conditional on the larger-population
  activity $\yA$.
\end{enumerate*}
Without this approximation the latter conditional frequency can differ from
$\yG_{\ya\yA}$ -- and this is expected if the number of bins $T$ is not much
larger than the number of possible joint possibilities, roughly
$n\times N$.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-05T10:25:02+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand*{\yr}{\bm{r}}

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{P_{\text{me}}}
\newcommand*{\pxx}{p}
\newcommand*{\pxxx}{P}
\newcommand*{\Nd}{\textrm{N}}
\newcommand*{\ycmm}{C}
\newcommand*{\ycm}{\mathte{\ycmm}}
\newcommand*{\ycs}{\bm{\sigma}}
\newcommand*{\yL}{L}

\newcommand*{\yla}{\bm{\lambda}}
\newcommand*{\yJ}{\bm{J}}
%\newcommand*{\yd}{\bm{\varDelta}}
\newcommand*{\ydi}{\varDelta}

@@

\section{Assumptions and approximations: mathematical sketch}
\label{sec:inference_prob}

We give here a sketch of the mathematical formulae behind the assumptions
and approximations discussed in \sect~\ref{sec:assumptions}. We indicate
them all collectively with $\yI$.

The first assumption, of infinite exchangeability
\citep[\chap~4]{bernardoetal1994_r2000}{dawid2013} of the probability
distribution $\pf[(\yA_{1},\dotsc, \yA_{T}) \| \yI]$ for the sequence of
activities, implies the following mixture form by de~Finetti's theorem
\citep{definetti1930,hewittetal1955}:
\begin{equation}
  \label{eq:mixture_from_definetti}
%  \exp[-T\;\sh(\yF;\ynu)] \quad
%  \pxxx[(\yA_{1},\dotsc,\yA_{T})] \propto
  \pf[(\yA_{1},\dotsc, \yA_{T}) \| \yI]
  = \int\!\!\!\di\ynu\;\pf(\ynu \|\yI)\,\prod_{\yA=1}^{N}{\nu_{\yA}}^{T\,\yFF_{\yA}},
\end{equation}
where $\ynu\defd(\nu_{\yA})$ is the long-run frequency distribution of
activities in an imaginary continuation or repetition of the recording, and
$\pf(\ynu \|\yI)\;\di\ynu$ our belief distribution about it.

The second assumption concerns the latter degree distribution. It has one
of these expressions:
\begin{subequations}\label{eq:maxent_prior}
  \begin{equation}
    \label{eq:entropic_prior}
    \pf(\ynu \| \yI) \propto  \exp[-\yL\;\sh(\yF;\yr)]
    \approx \binom{L}{L\nu_{0}, \dotsc, L\nu_{N}}\;
    \prod_{\yA}{r_{\yA}}^{L\nu_{\yA}}  
  \end{equation}
  where $\yr\defd(r_{\yA})$ is a reference distribution and the expression
  in larger parentheses is a multinomial coefficient, or
  \begin{equation}
    \label{eq:suff_prior}
    \pf(\ynu \| \yI) \propto
    \int\!\!\!\di\yla\;\pf(\yla \| \yI)\;
    \prod_{\yA}\delt\Biggl\{\ynu_{\yA}-
    \frac{r_{\yA}}{Z(\yla)}\,\exp\Bigl[
    \tsum_{m}\lambda_{m}\tbinom{\yA}{m}/\tbinom{N}{m}
    \Bigr]\Biggr\},
  \end{equation}
\end{subequations}
where $\yla\defd(\lambda_{m})$ are Lagrange multipliers and $Z(\yla)$ a
normalization factor for the exponential distribution within the delta.
This is a formal expression saying that our belief about $\ynu$ is an
exponential parametric model for which the normalized factorial moments are
informationally sufficient.

The third assumption corresponds to the formula
\begin{multline}
  \label{eq:assume_sampled_anew}
  \pf[(\ya_{1},\dotsc,\ya_{T}) \| (\yA_{1}, \dotsc, \yA_{T}), \yI ]
  ={}\\ \prod_{t=1}^{T} \pf(\ya_{t} \| \yA_{t}, \yI)
  = \prod_{t=1}^{T} \yG_{\ya_{t}\yA_{t}}
  \equiv \prod_{\ya,\yA}{\yG_{\ya_{t}\yA_{t}}}^{T\,J_{\ya\yA}},
\end{multline}
where $J_{\ya\yA}$ is the joint frequency distribution of the activity
pairs $(\ya,\yA)$, so that its marginals are the frequency distributions
$\yf$ and $\yF$. As discussed in \sect~\ref{sec:assumptions} this formula
follows from either of two beliefs: that all subsets of $\yA$ neurons in
the larger population are always equally likely to give rise to total
activity $\yA$, or that a new sampling is done at every time bin.

It can be proved that the formulae above lead to the following belief
distribution for the joint frequency distribution $\yJ \defd (J_{\ya\yA})$
conditional on $\yf$:
\begin{multline}
  \label{eq:prob_joint_frequencies}
  \pf(\yJ \| \yf, \yI) \propto
                         \int\!\!\!\di\ynu\;\pf(\ynu\|\yI)\;
                         \binom{T}{TJ_{0 0},\dotsc,TJ_{n N}}\,
                         \prod_{\ya,\yA}\bigl(\yG_{\ya\yA}\;\nu_{\yA}\bigr)^{T\,J_{\ya\yA}}
  \approx{}  \\
\shoveright{\int\!\!\!\di\ynu\;\pf(\ynu\|\yI)\;
                         \exp\biggl(
-T \tsum_{\ya,\yA}J_{\ya\yA}\ln\frac{J_{\ya\yA}}{\yG_{\ya\yA}\;\nu_{\yA}}
                         \biggr)}
  \\
  \text{with}\quad \tsum_{\yA}J_{\ya\yA}=\yff_{\ya},
\end{multline}
where $\pf(\ynu\|\yI)$ is either of the
distributions~\eqref{eq:maxent_prior}. From this belief distribution we can
find the one for $\yFF_{\yA}\equiv \sum_{\ya}J_{\ya\yA}$ by
marginalization. Within the exponential we can recognize the relative
entropy of $\yJ$ with respect to $(\yG_{\ya\yA}\;\nu_{\yA})$, which is the
reason why these two joint distributions must be equal if $T$ tends to
infinity.
